Immutable Source-Mapped Abstract Syntax Tree:\ \, \\ A Design Pattern for Refactoring Engine APIs

### 2017-10-03 paper.tex
Made forces more bulleted

To perform this refactoring, the conditional expression must be reversed in
meaning and the contents of the \textit{if} and \textit{else} branches must be
interchanged. Changing the condition requires removing the \textit{not} and the 
redundant parentheses. The former is simple enough, but the latter requires 
knowledge of matching parentheses and, in more complex cases, knowledge of operator
precedence. Interchanging the branches similarly requires 
knowledge of the matching braces, the location of the \textit{if}, and presence
or absence of the dangling 
\textit{else}.  In essence, knowledge of the program's syntax is 
required.

@@

@@

\item The API must expose fine-grained details about the syntax of a program.
For example, the preceding example requires being able to determine the
exact location of parentheses \ttt{(} \ttt{)}.

\item Many refactorings are initiated only after the user selects a range of
source code in a text editor.  Therefore, the API must allow clients to
determine what syntactic construct(s) are selected.

\item Clients must be able to use the API to manipulate source code in a way
that is minimally invasive: refactorings that only affect a small, isolated
region of the source code should not change any source code outside that
region.  This includes preserving the program's comments and
formatting~\cite{sommerlad08retaining}.

\item The API must follow general principles of good API design, including
being easy to learn and difficult to misuse.

\item The API should be usable when building refactorings for large code bases.
Automated refactorings are most useful for maintaining such code.

@@

\subsection{Abstract Syntax Trees}
\label{ss:asts}

Figure~\ref{fig:jdt-ast} shows a visualization of an AST for the following
program:
\vspace*{.5em}

\noindent \ttt{public class Example \{ \\
\hspace*{.2em} public int FIVE = 5; \\
\}
}

\vspace*{.5em}
Notice how deeper levels of the tree describe progressively more fine-grained
syntactic structures.  The root node represents the entire Java source file.
The highlighted node---a \textit{VariableDeclarationFragment}---represents
\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
since these tokens (like keywords, identifiers, and numeric literals) represent
the smallest syntactic units in the language.

@@

@@

To illustrate the first requirement, consider an example from Fortran.  The
Fortran language contains a number of input/output statements.  Two of these
are \textsc{print} and \textsc{write}: the statements \ttt{print *, "X"} and
\ttt{write (*,*) "X"} can be used interchangeably to print the string \ttt{X}
to standard output.  In general, the statement \ttt{print} \textit{fmt}\ttt{,}
\textit{expr} is equivalent to the statement \ttt{write (*,
}\textit{fmt}\ttt{)} \textit{expr}.  Most compilers type check and translate
them identically.  So, a compiler may choose to represent \textsc{print}
statements as \textsc{write} statements in its AST.  The GNU Fortran compiler
does this, for example.  However, onsider a program that uses \textsc{print} 
statements to write to standard output, but the programmer later needs to 
use \textsc{write}
statements to write to a file instead. Then the ability to convert \textsc{print}
statements to \textsc{write} statements actually makes for a useful
refactoring. To implement this refactoring, a
refactoring tool needs to be able to distinguish \textsc{print} statements from
\textsc{write} statements---a difficult task if they are represented
identically in the AST.

@@

\subsection{Mutability and ASTs}
\label{ss:manip}

There are two main variants of abstract syntax tree:
mutable and immutable. A mutable tree has nodes that can be easily added, 
changed, or deleted from the tree. These trees are often used internally by
compilers. For example, a compiler might determine that
\ttt{a~=~b*1} can be simplified to \ttt{a~=~b}, and then change the AST 
accordingly.

At first, this seems like a reasonable mechanism for refactoring tools to use
as well.  To remove a field declaration from a program, simply delete its
\textit{FieldDeclaration} node from the AST (along with all its children).  To
change the name of \ttt{FIVE}, change the contents of its
\textit{SimpleNameNode} in the AST.  Of course, some mechanism is needed to
prettyprint the modified AST, so these changes can be translated back into
source code, but that is relatively straightforward to implement.

Mutable ASTs are simple to understand and allow code for specific refactorings
to be very short. Performing refactorings on source code is conceptually
identical to changing the AST itself. This can greatly lower the bar for 
designing refactorings. However, mutable ASTs introduce a host of problems when
used in a large API:

\section{Solution}

\subsection{Immutability}
As seen in Section 3.4, mutability introduces a host of problems to ASTs. Mutable ASTs were used in the original Smalltalk Refactoring
Browser~\cite{roberts97refactoring}, as well as in
CRefactory~\cite{garrido05program} and Photran~\cite{sle2008}.  Many compilers
use mutable ASTs internally.  However, these tools have one thing in common:
the mutable AST is not accessible via any public API.  It is intended for
internal use only, by those tools' developers who should be better equipped to handle all of
the caveats.

@@

As discussed in Section 3.3, it must be possible to map nodes in the AST to
locations in the source code, and vice versa. The easiest way to map AST nodes to source locations is to store source
location information in the AST nodes themselves.  Again, consider the JDT AST
shown in Figure~\ref{fig:jdt-ast}, and note the numbers in square brackets.
The highlighted node is displayed as ``VariableDeclarationFragment [35, 8]'':
this AST node corresponds to the eight-character string beginning at offset 35
(\ttt{FIVE = 5}).  Its child node corresponding to the identifier \ttt{FIVE} is
attributed with the offset-length pair [35, 4], while the numeric literal
\ttt{5}'s source location is [42, 1].

@@

>>>>

To perform this refactoring, the conditional expression must be reversed in
meaning and the contents of the \textit{if} and \textit{else} branches must be
interchanged. Changing the condition requires removing the \textit{not} and the 
redundant parentheses. The former is simple enough, but the latter requires 
knowledge of matching parentheses and, in more complex cases, knowledge of operator

precedence. Interchanging the branches similarly requires 
knowledge of the matching braces, the location of the \textit{if}, and presence
or absence of the dangling 
\textit{else}.  In essence, knowledge of the program's syntax is 
required.

@@

\section{Background: Abstract Syntax Trees}
\label{ss:asts}

Figure~\ref{fig:jdt-ast} shows a visualization of an AST for the following
program:

\begin{center}
\vspace*{1em}
\begin{tabular}{ccc}
\begin{minipage}{2.25in}
\noindent \ttt{public class Example \{ \\
\hspace*{1.5em} public int FIVE = 5; \\
\}
}
\end{minipage}
\end{tabular}
\vspace*{1em}
\end{center}

Notice how deeper levels of the tree describe progressively more fine-grained
syntactic structures.  The root node represents the entire Java source file.
The highlighted node---a \textit{VariableDeclarationFragment}---represents
\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
since these tokens (like keywords, identifiers, and numeric literals) represent
the smallest syntactic units in the language.

There are two main variants of abstract syntax tree:
mutable and immutable. A mutable tree has nodes that can be easily added, 
changed, or deleted from the tree. 
These trees are often used internally by

compilers. For example, a compiler might determine that
\ttt{a~=~b*1} can be simplified to \ttt{a~=~b}, and then change

the AST accordingly. Immutable trees, on the other hand, cannot be changed via
function calls. These can be used for read-only purposes or operated on in a
functional manner.

@@

    \item \textit{The API must expose fine-grained details about the syntax of a program.}
For example, the preceding example requires being able to determine the
exact location of parentheses \ttt{(} \ttt{)}.

\item \textit{Many refactorings are initiated only after the user selects a range of
source code in a text editor.  Therefore, the API must allow clients to
        determine what syntactic construct(s) are selected.}

    \item \textit{Clients must be able to use the API to manipulate source code in a way
        that is minimally invasive.} Refactorings that only affect a small, isolated
region of the source code should not change any source code outside that
region.  This includes preserving the program's comments and
formatting~\cite{sommerlad08retaining}.

\item \textit{The API must follow general principles of good API design, including
    being easy to learn and difficult to misuse.}

\item \textit{The API should be usable when building refactorings for large code bases.}
Automated refactorings are most useful for maintaining such code.

@@

\subsection{AST Forces}
\label{ss:manip}

Since mutable ASTs are used by compilers, they seem like a reasonable mechanism for refactoring tools to use
as well.  To remove a field declaration from a program, simply delete its
\textit{FieldDeclaration} node from the AST (along with all its children).  To
change the name of \ttt{FIVE}, change the contents of its
\textit{SimpleNameNode} in the AST.  Of course, some mechanism is needed to
prettyprint the modified AST, so these changes can be translated back into
source code, but that is relatively straightforward to implement.

Mutable ASTs have several benefits:
\begin{itemize}
    \item \textit{Mutable ASTs are simple to understand.} Performing 
        refactorings on source code is conceptually identical to changing 
        the AST itself.
    \item \textit{Source code for specific refactorings can be very short.} 
        The code simply has to rearrange the AST.

However, mutability introduces several issues:

@@

Any candidate solution would have to satisfy the API requirements while avoiding
the problems associated with mutable ASTs.

\clearpage
\section{Solution}

@@

To illustrate the first requirement, consider an example from Fortran.  The
Fortran language contains a number of input/output statements.  Two of these
are \textsc{print} and \textsc{write}: the statements \ttt{print *, "X"} and
\ttt{write (*,*) "X"} can be used interchangeably to print the string \ttt{X}
to standard output.  In general, the statement \ttt{print} \textit{fmt}\ttt{,}
\textit{expr} is equivalent to the statement \ttt{write (*,
}\textit{fmt}\ttt{)} \textit{expr}.  Most compilers type check and translate
them identically.  So, a compiler may choose to represent \textsc{print}
statements as \textsc{write} statements in its AST.  The GNU Fortran compiler
does this, for example.  However, consider a program that uses \textsc{print} 
statements to write to standard output, but the programmer later needs to 
use \textsc{write}
statements to write to a file instead. Then the ability to convert \textsc{print}
statements to \textsc{write} statements actually makes for a useful
refactoring. To implement this refactoring, a
refactoring tool needs to be able to distinguish \textsc{print} statements from
\textsc{write} statements---a difficult task if they are represented
identically in the AST.

@@

@@

The easiest way to map AST nodes to source locations is to store source
location information in the AST nodes themselves.  Again, consider the JDT AST
shown in Figure~\ref{fig:jdt-ast}, and note the numbers in square brackets.
The highlighted node is displayed as ``VariableDeclarationFragment [35, 8]'':
this AST node corresponds to the eight-character string beginning at offset 35
(\ttt{FIVE = 5}).  Its child node corresponding to the identifier \ttt{FIVE} is
attributed with the offset-length pair [35, 4], while the numeric literal
\ttt{5}'s source location is [42, 1].

@@

\subsection{Immutability}

Mutable ASTs were used in the original Smalltalk Refactoring
Browser~\cite{roberts97refactoring}, as well as in
CRefactory~\cite{garrido05program} and Photran~\cite{sle2008}.  Many compilers
use mutable ASTs internally.  However, these tools have one thing in common:
the mutable AST is not accessible via any public API.  It is intended for
internal use only, by those tools' developers who should be better equipped to handle all of
the caveats.
